---
title: "01_data-download"
author: "Kate Sheridan"
date: "9/13/2022"
output: html_document
---

Reduced OBIS/GBIF dataset for reproducibility course.

Noting here that I am not using groundhog even though the checklist wants you to because I would have to set it up on multiple computers and change a lot of scripts that I'm currently actively working on for my thesis.

```{r setup, include=FALSE}

## Install anything you don't have!
#install.packages('robis')
#install.packages('rgbif')
#install.packages('worrms')
#install.packages('here')

# load libraries
library(tidyverse)
library(robis)
library(rgbif)
library(worrms)
library(taxize)
library(here)
```

This is a custom function I wrote to use worrms to extract taxonomy information for species using the worrms package. Names are extracted from the World Record of Marine Species (WoRMS), which is the most authoritative source for marine taxonomy. Using this call, I am also able to extract other information from the WoRMS that I can't with taxize; specifically "is_terrestrial" and "is_freshwater", which I use to filter out species that are only found in non-marine realms but are present in the data.

```{r function worrms}
## requires worrms to be loaded
# input is vector of species names
# uses wm_records_names to extract 

# right now this breaks if not-found
search_records_worms <- function(spnames) {
  search <- tibble()
  for (i in spnames) {
    print(paste0('searching for ', i))
    record <- wm_records_names(i, marine_only = FALSE)
    message('done')
    search <- append(search, record)
  }
  names(search) <- spnames
  search_output <- map_dfr(.x = search, ~ data.frame(.x), .id = 'query') %>%
    janitor::clean_names() %>%
    select(!(c(url, taxon_rank_id, citation, lsid, modified)))
  return(search_output)
}
```


# GBIF/OBIS pulls

## Search parameters for gbif/obis

The taxa searched are families from two classes identified in recent seagrass nearshore eDNA samples amplified for COI. Both had the highest alpha diversity by two different metrics:
Bivalves had the highest genetic diversity (unique ASVs in dataset)
Polychaetes had the highest species diversity (unique species assignments)

Final analyses will be run on a different eDNA dataset, amplified for a different marker (12S), and most likely fish only, so this selection maintains 'researcher degrees of freedom'.
```{r parameters}
family_list <- c(# Bivalves
              "Veneridae", "Myidae", "Glycymerididae",
              "Lasaeidae", "Mactridae", "Tellinidae",
                # Polychaetes
              "Polynoidae", "Glyceridae", "Sabellidae",  
              "Goniadidae", "Capitellidae", "Syllidae",
              "Nereididae", "Orbiniidae", "Phyllodocidae", 
              "Sigalionidae")



# Bounding box for Pacific coast, Alaska to equator
# close to nearshore
westcoast_bb <- "POLYGON((-168 69, -168.7 64.8, -166 63.5, -168.3 60, -162.2 57, -160 54.4, -148 57, -138 58, -133 52, -125.6 48, -126 42.7, -123 34, -117 30, -107.5 19.3, -94 14, -87.3 10.5, -81.5 6.6, -79.4 2.3, -82.4 -1, -80 -1, -76.8 3.7, -77.25 7.25, -78.56 9.1, -79.5 9.2, -81.3 8.1, -85.71 11.23, -87.27 13.5, -91.34 14.38, -94.4 16.6, -96.6 15.9, -105 20.2, -104.9 21.8, -112.8 31.4, -120 34.75, -121.8 37.2, -121.6 38, -122.75 38.2, -124 40.4, -123.8 41.6, -124.3 42.85, -123 46, -122.3 47, -122 47.6, -122.4 49.4, -126.6 51.5, -129.6 54.6, -130.85 56.1, -133.6 58.4, -135.6 59.5, -137.5 58.9, -139.5 60, -143.9 60.2, -146 61.2, -149.7 61.6, -151.75 61.1, -157.6 59, -161.4 59.3, -163.8 60.6, -163.8 60, -165.4 61.7, -162.2 63.3, -160.7 63.5, -160.7 65, -159.5 66.5, -165.3 68.5, -156.7 71.1, -168 69))"
```

### GBIF

DOI: 10.15468/dl.43cw9n
  Citation:
  GBIF Occurrence Download https://doi.org/10.15468/dl.43cw9n Accessed from R via rgbif (https://github.com/ropensci/rgbif) on 2022-09-07
```{r gbif records}
gbif_taxon_keys <-
  family_list %>%
  # match names to the GBIF backbone to get taxonkeys
  get_gbifid_(method = "backbone") %>% 
  # add original name back into data.frame
  imap(~ .x %>% mutate(original_sciname = .y)) %>% 
  # combine all data.frames into one
  bind_rows() %>% 
  # get only accepted and matched names
  filter(matchtype == "EXACT" & status == "ACCEPTED") %>% 
  # avoid homonyms with plants
  filter(kingdom == "Animalia") %>%
  # this search was at the family level, get rid of generic matches
  filter(rank == "family")

# download data 
# redact for ldp version
user <- "your_username"
pwd <- "your_password"
email <- "your_email"

#specific taxa within bounding box
occ_download(pred_in("taxonKey", unique(gbif_taxon_keys$usagekey)),
             pred("hasCoordinate", TRUE), 
             pred_within(westcoast_bb), 
             format = "SIMPLE_CSV", 
             user = user, pwd = pwd, email = email)

#then go to gbif to download the file, 
#or get it with the occ_download_get function using my key
gbif_dat <- occ_download_get('0006322-220831081235567') %>%
    occ_download_import()
```
GBIF records need cleaning, we will remove records that are problematic.
Also taxonomy should be standardized to WoRMS for integration with other datasets, and because WoRMS is the standard for marine taxonomy.

```{r gbif-clean}
# remove records with low coordinate precision
 # check
#hist(gbif_dat$coordinateUncertaintyInMeters/1000, breaks = 30)
gbif_cl <- gbif_dat %>% 
  filter(coordinateUncertaintyInMeters/1000 <= 100|
           is.na(coordinateUncertaintyInMeters)) %>%
# remove unsuitable data sources, especially fossils
  filter(basisOfRecord == "HUMAN_OBSERVATION" | 
           basisOfRecord == "OBSERVATION" | 
           basisOfRecord == "PRESERVED_SPECIMEN" | 
           basisOfRecord == "MATERIAL_SAMPLE" | is.na(basisOfRecord)) %>%
  # remove absences
  filter(occurrenceStatus == "PRESENT") %>%
  # remove 0s; records usually have an issue
  filter(individualCount > 0 | is.na(individualCount)) %>%
  # remove records before second world war
  # change in recording practices and land use around this time
  # records prior tend to be less reliable also
  filter(year > 1944) %>%
  # clean gbif sciname columm
  mutate(gbif_sciname = ifelse(species == "", genus, species)) %>%
  # remove anything without generic or species level id
  filter(!(gbif_sciname == "")) %>%
  mutate(taxonRank = tolower(taxonRank)) %>%
    # filter to only useful columns
  # removing higher taxonomy
  select(gbifID, gbif_sciname, taxonRank, scientificName, taxonKey,
         year, countryCode, stateProvince, locality,
         decimalLatitude, decimalLongitude, depth
         )

```

Taxonomy cleaning; for gbif 
```{r}
## Taxonomy cleaning
gbifids <- gbif_cl %>%
  filter(taxonRank %in% c("genus", "species",
                          "subspecies", "unranked")) %>%
  # first extract species gbifids from search
  select(taxonKey, gbif_sciname, taxonRank) %>%
  distinct(gbif_sciname, .keep_all = TRUE) %>%
  # worms is having an issue with curl 
  # searches needs + instead of whitespace
  mutate(wormsquery = gsub(" ", "+", gbif_sciname))

# get worms ids to update taxonomy
## let it find non marine and unaccepted names!
## this can take a while
gbif_worms <- get_wormsid_(sci_com = gbifids$wormsquery, 
                           marine_only = FALSE,
                           accepted = FALSE)

# get_wormsid_ gives back null and length 0 elements
## first remove with compact and discard then map_dfr
gbif_worms2 <- gbif_worms %>% 
  compact() %>%
  discard( ~ nrow(.x) == 0) %>%
  map_dfr( ~ data.frame(.x), .id = 'wormsquery') %>%
  left_join(gbifids) %>%
  relocate(gbif_sciname) %>%
  select(!(wormsquery)) %>%
  filter(!(!(gbif_sciname == scientificname))) %>%
  rename(worms_aphiaid = AphiaID,
        worms_sciname = scientificname) %>%
  distinct()

# custom function to get higher taxonomy and additional data
gbif_worms_records <- search_records_worms(gbif_worms2$worms_sciname)

# clean the output
gbif_worms_records2 <- gbif_worms_records %>%
  # sometimes gets plants
  filter(kingdom == 'Animalia') %>%
  # get rid of non target taxa
  filter(class %in% c('Polychaeta', 'Bivalvia')) %>%
  # make freshwater only and terrestrial only columns for filtering
  mutate(freshwater_only = ifelse((is_marine %in% c(0, NA) ==TRUE) & (is_brackish %in% c(0,NA) == TRUE) & (is_freshwater == 1),  "1", "0")) %>%
    mutate(terrestrial_only = ifelse((is_marine %in% c(0, NA) ==TRUE) & (is_brackish %in% c(0,NA) == TRUE) & (is_terrestrial == 1),  "1", "0")) %>%
  distinct() %>% 
# rename to format
  rename(gbif_sciname = query,
         worms_aphiaid = valid_aphia_id,
         original_aphiaid =aphia_id,
         worms_match = scientificname,
         worms_sciname = valid_name
         ) %>%
  relocate(c(worms_sciname, worms_aphiaid), .before = original_aphiaid) %>%
  # rank lowercase
  mutate(rank = tolower(rank)) %>%
  # filter out junior homonyms
  group_by(gbif_sciname) %>%
  mutate(n_found = n()) %>%
  ungroup() %>%
  filter(!(n_found > 1 & status != 'accepted')) %>%
    # remove unneeded columns
  select(!(c(authority, parent_name_usage_id, is_freshwater, 
             is_terrestrial, is_marine, is_brackish, 
             is_extinct, match_type, n_found, kingdom)))
  
```


Merge taxonomy back to the cleaned gbif records: remove all the gbif taxonomy to replace with WoRMS but retain gbif IDs.
```{r gbif merge taxonomy}
gbif_cl_tax <- gbifids %>%
  select(!(wormsquery)) %>%
  left_join(gbif_worms_records2) %>%
  # remove records with bad taxonomy
  filter(!(is.na(worms_sciname))) %>%
  # rely on worms to select only genus or 
  ## species level ID rather than gbif
  filter(rank %in% c('genus', 'species')) %>%
  # join with locality data
  right_join(gbif_cl) %>%
  #remove all records without worms IDs
  filter(!(is.na(worms_sciname))) %>%
  # redundant column
  select(!(taxonRank)) %>%
  rename(gbif_taxonkey = taxonKey,
         gbif_verbatim = scientificName) %>% 
  mutate(terrestrial_only = as.numeric(terrestrial_only))


# save to csv
write_csv(gbif_cl_tax, here('data', 'rawdata',
                            '20220909_gbif_bivalve-polychaete-subset.csv'))

#TA's suggestaion; incorporate later
#write.csv(object, file =paste0("name",Sys.Date(), ".csv"))

  ## identify taxonomic problems
problems <- gbifids %>%
  select(!(wormsquery)) %>%
  left_join(gbif_worms_records2)%>% 
  filter(is.na(worms_sciname))
# save problem list to review later
#write_csv(problems, here('processeddata', 'bigdata',
#                         '20220909_gbif_subset-problems.csv'))

```



### OBIS

```{r obis records}
obis_taxon_keys <-
  family_list %>%
  # match names to the WoRMS backbone to get taxonkeys
  get_wormsid_(method = "backbone") %>% 
  # add original name back into data.frame
  imap(~ .x %>% mutate(original_sciname = .y)) %>% 
  # combine all data.frames into one
  bind_rows() 

# download directly (its fast)
obis_dat <- occurrence(taxonid = obis_taxon_keys$AphiaID, 
                      geometry = westcoast_bb)

```


```{r obis-clean}
# check fields to determine what to remove
table(obis_dat$eventDate)
table(obis_dat$basisOfRecord)
table(obis_cl$flags)

obis_cl <- obis_dat %>%
  #remove records without dates
  filter(!is.na(eventDate)) %>%
  # also invalid/bad dates
  filter(!(eventDate %in% c('0000-00-00', '0000-02-00',
                          '0000-10-00'))) %>%
  # remove unsuitable data sources
  filter(basisOfRecord == "Human observation" | basisOfRecord == "HumanObservation" | basisOfRecord == "PreservedSpecimen" | basisOfRecord == "MaterialSample" | basisOfRecord == 'LivingSpecimen' | is.na(basisOfRecord)) %>%
  # remove records on land
  filter(!(grepl('ON_LAND', flags))) %>%
  # identify terrestrial-only species
  ## there were none, remove if there are some in future datasets
  mutate(terrestrial_only = ifelse((terrestrial == "TRUE" & brackish == "FALSE" & marine =="FALSE"), 1, 0)) %>% 
  # select only columns we need
  select(scientificName, aphiaID, taxonRank, 
         phylum, class, subclass, order, family, genus,
         year, countryCode, stateProvince, county, locality,
         decimalLatitude, decimalLongitude, depth,
         taxonRemarks, originalScientificName, terrestrial_only) %>% 
  # recode + populate taxon rank so the column is usable
  mutate(taxonRank = tolower(taxonRank)) %>%
  mutate(taxonRank = recode(taxonRank, familia = 'family',
                            especie = 'species',
                            especies = 'species',
                            género = 'genus')) %>% 
  mutate(taxonRank = ifelse(is.na(genus), 'family', 
                            ifelse(str_detect(scientificName, " ") == FALSE &
                                     is.na(taxonRank), 'genus', 
                                   ifelse(str_detect(scientificName, " ") == TRUE &
                                     is.na(taxonRank), 'species', taxonRank)))) %>% 
  # remove family level IDs
  filter(!(taxonRank == 'family')) %>%
  # make columns match gbif clean
  rename(worms_sciname = scientificName,
         worms_aphiaid = aphiaID,
         verbatim_obis = originalScientificName,
         rank = taxonRank) %>% 
  mutate(year = as.numeric(year))
  

# save to csv
write_csv(obis_cl, here('data', 'rawdata',
                            '20220909_obis_bivalve-polychaete-subset.csv'))

```


## Merge gbif-obis

```{r}
# load in cleaned files if needed
#gbif_cl_tax <- read_csv(here('data', 'rawdata', '20220909_gbif_bivalve-polychaete-subset.csv'))
#obis_cl <- read_csv(here('data', 'rawdata', '20220909_obis_bivalve-polychaete-subset.csv'))

bigdat <- obis_cl %>% 
  full_join(gbif_cl_tax) %>%
  # remove putative duplicates;
  # we're also not looking at seasonality so one occurrence per locality per year is sufficient.
  distinct(worms_sciname, year, decimalLongitude, decimalLatitude, .keep_all = TRUE) %>%
  #standardize names (sorry darwincore)
  janitor::clean_names()

# save to csv
write_csv(bigdat, here('data', 'rawdata',
                            '20220909_obis-gbif_bivalve-polychaete-subset.csv'))

```


