---
title: "03_mCART"
author: "Kate Sheridan"
date: "9/22/2022"
output: html_document
---

# IMPORTANT NOTE
Key package mvpart is not currently available on cran and needs to be installed from the archives. Use devtools to install!

# Overview

This script does the following:
1) loads in only the class needed, identified to species, within a buffer around the study range
2) Generate range-through data
3) Use range-through to map species presence in 1-degree bins across their range
4) Generate dissimilarity matrix from presence data
5) Visualize with nMDS
6) Run mCART (Multivariate Classification and Regression Tree) on dissimilarity matrix vs latitude to determine splitting between regions; note that in this version of the analysis no environmental data has been provided to the analysis, clusters are formed based only on community P/A data. next iteration will include environmental variables. (No time to get those also tidied up in time for the project!)
7) export results in a plottable format

# Setup

```{r setup, include=FALSE}
# mvpart needs to be installed from CRAN archives
#devtools::install_github("cran/mvpart")
library(mvpart)

#install.packages('data.tree')
library(data.tree)

library(tidyverse)
library(stringr)
library(ggplot2)
library(vegan)
library(ape)
library(here)

```

```{r load-in}
# only polychaetes a bit outside study range
poly <- read_csv(here('data', 'rawdata',
                            '20220909_obis-gbif_bivalve-polychaete-subset.csv')) %>%
  # only polychaetes ID to species
  filter(class == "Polychaeta") %>%
  filter(rank == 'species') %>%
  # within latitude requirements
  filter(decimal_latitude > 25) %>%
  filter(decimal_latitude < 65)

```

# Format data for analysis

## Range-through

Following Fenberg et al 2015 and 2019, I generate my matrix based on the minimum and maximum latitude of species occurrence, rather than Presence-Absence within each 1-degree bin. 
At this point I filter to only the species found within our study range: 455 Polychaete species

```{r rangethrough}
# generate min and max ranges
poly_rt <- poly %>%
  group_by(worms_sciname) %>%
  summarise(min_lat = min(decimal_latitude), max_lat = max(decimal_latitude)) %>%
  # remove species that are unique to outside study range
  filter(max_lat >= 30 & max_lat <= 63) %>%
  rownames_to_column('index')

# write if needed
#write.csv(poly_rt, here('data', 'processeddata', '20220926_range-through_polychaete.csv'))

# make bins
# 32 1 degree latitude bins from 30-62N
bins <- data.frame(min = seq(30:61) + 29, max = seq(30:61) + 30) %>%
  rownames_to_column('bin_index')

#define binning function
## if the minimum latitude and maximum latitude permit the species to be in the bin,
## presence, else, absence.
bin_fun <- function(min_lat, max_lat, min, max){
  col <- ifelse((min_lat <= max & max_lat >= min), 1, 0)
  x <- as.data.frame(col)
}

# taxa P/A for bins
# Use purrr to run function on all species and bind resulting columns
poly_bin <- map2_dfc(poly_rt$min_lat, poly_rt$max_lat, bins$min, bins$max, .f = bin_fun)

# fix up to be usable
poly_bin <- poly_bin %>%
  # column for 'which bin'
  mutate(bin_lat = seq(30:61) + 29) %>%
  relocate(bin_lat) %>%
  # remove extra letters in names
  rename_with(., ~ str_remove_all(., 'col\\.\\.\\.')) %>%
  pivot_longer(-c("bin_lat"), names_to = 'index') %>%
  #species names into dataframe
  left_join(poly_rt) %>%
  select(!(c(index, min_lat, max_lat))) %>%
  #format into site x species P/A matrix
  pivot_wider(id_cols = bin_lat, names_from = worms_sciname) %>%
  # future analysis needs matrix only with rownames
  column_to_rownames('bin_lat')

```




## nMDS

Ordination for community data
Jaccard dissimilarity for P/A data
Extract plot

```{r nmds}
#run the ordination
poly_nmds <- poly_bin %>%
  #run ordination
  metaMDS(.,
          distance = "jaccard",
          k = 2,
          trymax = 5000)


# check stress
stressplot(poly_nmds)

#extract from NMDS
poly_nmds_scores <- as.data.frame(poly_nmds$points) %>%
  rownames_to_column(var = "lat_bin") 

# keeping the ordination simple
# we can see the latitudinal pattern
ggplot(poly_nmds_scores) +
  geom_point(aes(x = MDS1, y = MDS2, col = lat_bin)) +
  theme_bw()


ggsave(here('output', 'clustering', '20220926_nmds_polychaetes.png'))
```



## mCART

Use vegan to generate distance matrix using Jaccard for P/A data.
mvpart for mCART. 1se recommended for parsimony.

```{r}
# Jaccard distance
poly_jaccard <- vegdist(poly_bin, method = 'jaccard', binary = T)

poly_mvpart <- mvpart(as.matrix(poly_jaccard) ~ min,
                      # distance matrix vs latitude bins
                      data = bins,
                      # select minimum tree
                      ## within one standard error
                      xv = '1se',
                      # cross validation settings
                      xval = nrow(bins),
                      xvmult = 100,
                      # also plot PCA
                      pca = T,
                      # also plot rsquare plot
                      rsq = T
                      )
summary(poly_mvpart)

# tear apart the object and reconstruct it
# because mvpart is deprecated we need to manually extract
## the useful parts of the output for full control over plotting

poly_mvpart_data <- as.data.frame(poly_mvpart$frame) %>%
  # these row names are not what 'where' refers to
  rownames_to_column('junk') %>%
  # THESE are the cluster numbers
  rownames_to_column('cluster') %>%
  mutate(cluster = as.factor(cluster)) %>%
  # don't need the useless numbers
  ## or yval2
  select(!(junk)) %>%
  select(!(yval2))


## match up cluster numbers to latitude bins
poly_bin_clust <- as.data.frame(poly_mvpart$where) %>%
  rename(cluster = `poly_mvpart$where`) %>%
  mutate(cluster = as.factor(cluster)) %>%
  rownames_to_column('bin_index') %>%
  left_join(bins) %>%
  right_join(poly_mvpart_data) %>%
  filter(!(is.na(bin_index))) %>%
  #  remove redundant columns
  select(!(c(ncompete, nsurrogate, var)))

write.csv(poly_bin_clust, file = here('data', 'processeddata', '20220926_mcart-cluster_polychaete.csv'))
  
# now extract tree information for plotting
# we need to cycle the data through packages to convert the information
# first data.tree, convert rpart object to data.tree object
poly_node <-as.Node(poly_mvpart)
# data.tree to convert data.tree to phylo for ape
poly_phylo <- as.phylo.Node(poly_node)
# ape to save tree
write.tree(poly_phylo, file = here('data', 'processeddata', '20220926_mcart-cluster_polychaete.tre'))

```






Export a png of the pca for now
```{r pca}
png(filename = here('output', 'clustering', '20220926_polychaete_pca.png'), type = 'quartz')
rpart.pca(poly_mvpart)
dev.off()
```

