---
title: "03b_biogeographic_breaks-bivalve"
author: "Kate Sheridan"
date: "9/24/2022"
output: html_document
---
# IMPORTANT NOTE
Key package mvpart is not currently available on cran and needs to be installed from the archives. Use devtools to install!


# Overview
This script is identical to the polychaete script but for bivalves, see 03a for more details and annotations.


# Setup
```{r setup, include=FALSE}
## install mvpart from the cran archive with devtools
#devtools::install_github("cran/mvpart")
library(mvpart)

## for converting to useful formats
#install.packages('data.tree')
#install.packages('vegan')
#install.packages('ape')
#install.packages('stringr')

library(tidyverse)
library(stringr)
library(ggplot2)
library(data.tree)
library(vegan)
library(ape)
library(here)
```

```{r load-in}
# only polychaetes a bit outside study range
biv <- read_csv(here('data', 'rawdata',
                            '20220909_obis-gbif_bivalve-polychaete-subset.csv')) %>%
  filter(class == "Bivalvia") %>%
  filter(rank == 'species') %>%
  filter(decimal_latitude > 25) %>%
  filter(decimal_latitude < 65)

```

# Format data for analysis

## Range-through

Unique to range: 144 Bivalve species
```{r rangethrough}
# generate min and max ranges
biv_rt <- biv %>%
  group_by(worms_sciname) %>%
  summarise(min_lat = min(decimal_latitude), max_lat = max(decimal_latitude)) %>%
  # remove species that are unique to outside study range
  filter(max_lat >= 30 & max_lat <= 63) %>%
  rownames_to_column('index')

# write if needed
#write.csv(biv_rt, here('data', 'processeddata', '20220926_range-through_bivalve.csv'))

# make bins
# 32 1 degree latitude bins from 30-62N
bins <- data.frame(min = seq(30:61) + 29, max = seq(30:61) + 30) %>%
  rownames_to_column('bin_index')

#define binning function
bin_fun <- function(min_lat, max_lat, min, max){
  col <- ifelse((min_lat <= max & max_lat >= min), 1, 0)
  x <- as.data.frame(col)
}

# taxa P/A for bins
biv_bin <- map2_dfc(biv_rt$min_lat, biv_rt$max_lat, bins$min, bins$max, .f = bin_fun)

# fix up to be usable
biv_bin <- biv_bin %>%
  # column for which bin
  mutate(bin_lat = seq(30:61) +29) %>%
  relocate(bin_lat) %>%
  # remove extra letters in names
  rename_with(., ~ str_remove_all(., 'col\\.\\.\\.')) %>%
  pivot_longer(-c("bin_lat"), names_to = 'index') %>%
  #species names into dataframe
  left_join(biv_rt) %>%
  select(!(c(index, min_lat, max_lat))) %>%
  #format into site x species P/A matrix
  pivot_wider(id_cols = bin_lat, names_from = worms_sciname) %>%
  column_to_rownames('bin_lat')

```


## nMDS

```{r}
#run the ordination
biv_nmds <- biv_bin %>%
  #run ordination
  metaMDS(.,
          distance = "jaccard",
          k = 2,
          trymax = 5000)

# check stress
stressplot(biv_nmds)

#extract from NMDS
biv_nmds_scores <- as.data.frame(biv_nmds$points) %>%
  rownames_to_column(var = "lat_bin") 

# keeping the ordination simple
# we can see the latitudinal pattern
ggplot(biv_nmds_scores) +
  geom_point(aes(x = MDS1, y = MDS2, col = lat_bin)) +
  theme_bw()

ggsave(here('output', 'clustering', '20220926_nmds_bivalves.png'))
```

# mCART
```{r mcart}
# Jaccard distance
biv_jaccard <- vegdist(biv_bin, method = 'jaccard', binary = T)

biv_mvpart <- mvpart(as.matrix(biv_jaccard) ~ min,
                      # distance matrix vs latitude bins
                      data = bins,
                      # select minimum tree
                      ## within one standard error
                      xv = '1se',
                      # cross validation settings
                      xval = nrow(bins),
                      xvmult = 50,
                      # also plot PCA
                      pca = T,
                     rsq = T
                      )

# tear apart the object and reconstruct it
# because mvpart is deprecated we need to manually extract
## the useful parts of the output for full control over plotting

biv_mvpart_data <- as.data.frame(biv_mvpart$frame) %>%
  # these row names are not what 'where' refers to
  rownames_to_column('junk') %>%
  # THESE are the cluster numbers
  rownames_to_column('cluster') %>%
  mutate(cluster = as.factor(cluster)) %>%
  # don't need the useless numbers
  ## or yval2
  select(!(junk)) %>%
  select(!(yval2))


## match up cluster numbers to latitude bins
biv_bin_clust <- as.data.frame(biv_mvpart$where) %>%
  rename(cluster = `biv_mvpart$where`) %>%
  mutate(cluster = as.factor(cluster)) %>%
  rownames_to_column('bin_index') %>%
  left_join(bins) %>%
  right_join(biv_mvpart_data) %>%
  filter(!(is.na(bin_index))) %>%
  #  remove redundant columns
  select(!(c(ncompete, nsurrogate, var)))

write.csv(biv_bin_clust, file = here('data', 'processeddata', '20220926_mcart-cluster_bivalves.csv'))
  
# now extract tree information for plotting
# we need to cycle the data through packages to convert the information
# first data.tree, convert rpart object to data.tree object
biv_node <-as.Node(biv_mvpart)
# data.tree to convert data.tree to phylo for ape
biv_phylo <- as.phylo.Node(biv_node)
# ape to save tree
write.tree(biv_phylo, file = here('data', 'processeddata', '20220926_mcart-cluster_bivalves.tre'))
```


Export a png of the pca for now
```{r pca}
png(filename = here('output', 'clustering', '20220926_bivalve_pca.png'), type = 'quartz')
rpart.pca(biv_mvpart)
dev.off()
```
