---
title: "04_breaks-mapping"
author: "Kate Sheridan"
date: "9/27/2022"
output: html_document
---

This script will make neat maps from the clusters and visualize the ranges!

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
# i will use ggtree ultimately
#library(ggtree)
library(ggnewscale)
library(ape)
library(here)

# set up palettes for consistency
# 3,5,6,9,10,11 polychaetes
# 8,7,4,5,10,11 bivalves
### cluster equivalencies; P6 = B8, P5 = B7, P9 = B5, P3 = B4
biv_cluster_pal <- c('4' = '#d1d39b',  # neutral yellow
               '5' = '#bac9dd', #light blue 
               '7' = '#93bf6e', # yellow green 
              '8' = '#628e3d',  # darker green              
               '10' =  '#688ab5', #medium blue
              '11' = '#2a3e57' #darkest blue
              )

poly_cluster_pal <- c('3' = '#d1d39b',  # neutral yellow
               '5' = '#93bf6e', # yellow green 
               '6' = '#628e3d',  # darker green
               '9' = '#bac9dd', #light blue 
               '10' =  '#688ab5', #medium blue
              '11' = '#2a3e57' #darkest blue
              )
```

```{r load-in}
# range-through, cluster data, cluster tree, occurrences
poly_rt <- read_csv(here('data', 'processeddata', '20220926_range-through_polychaete.csv'))[-1]
poly_clust <- read_csv(here('data', 'processeddata', '20220926_mcart-cluster_polychaete.csv'))[-1]
poly_tree <- read.tree(here('data', 'processeddata', '20220926_mcart-cluster_polychaete.tre'))
# filter occurrences a priori
poly <- read_csv(here('data', 'rawdata',
                            '20220909_obis-gbif_bivalve-polychaete-subset.csv')) %>%
  filter(class == "Polychaeta") %>%
  # trim to study range
  filter(decimal_latitude > 30 & decimal_longitude < -116.4) %>%
  filter(decimal_latitude < 63 & decimal_longitude > -142.5) %>%
  filter(rank == 'species') %>%
  select(worms_sciname, decimal_latitude, decimal_longitude, 
         class, subclass, order, family, genus) %>%
    distinct()



biv_rt <-  read_csv(here('data', 'processeddata', '20220926_range-through_bivalve.csv'))[-1]
biv_clust <- read_csv(here('data', 'processeddata', '20220926_mcart-cluster_bivalves.csv'))[-1]
biv_tree <- read.tree(here('data', 'processeddata', '20220926_mcart-cluster_polychaete.tre'))
biv <- read_csv(here('data', 'rawdata',
                            '20220909_obis-gbif_bivalve-polychaete-subset.csv')) %>%
  filter(class == "Bivalvia") %>%
  # trim to study range
  filter(decimal_latitude > 30 & decimal_longitude < -116.4) %>%
  filter(decimal_latitude < 63 & decimal_longitude > -142.5) %>%
  filter(rank == 'species') %>%
  select(worms_sciname, decimal_latitude, decimal_longitude, 
         class, subclass, order, family, genus) %>%
  distinct()


# world map
world <- map_data("world")
```


### Visualize ranges

Use range-through data to visualize the minimum and maximum extents of ranges

Polychaete species have minimums and maximiums across the study area, but it seems like there are several common min and max bins. 

```{r range-plots}
### Polychaetes

# min and max for each species
## in points; one pair of points per species
## order by minimum latitude
ggplot(poly_rt) +
  geom_point(aes(x = reorder(index, min_lat), y = max_lat), 
             color = 'blue',
             alpha = .7) +
  geom_point(aes(x = reorder(index, min_lat), y = min_lat), 
             color = 'red',
             alpha = .4) +
  labs(title = 'Polychaete minimum and maximum occurrences', y = 'Latitude', x = 'Species') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_blank())

ggsave(here('output', 'sp_range', '20220926_minmax-dots_polychaetes.png'))


# in boxplots
ggplot(poly_rt) + 
  geom_boxplot(aes(y = max_lat, x = 1), 
               fill = 'blue', 
               alpha = .5) +
  geom_boxplot(aes(y = min_lat, x = 2), 
               fill = 'red', 
               alpha = .5) +
  labs(title = 'Polychaete minimum and maximum occurrences', y = 'Latitude', x = '') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid = element_blank())

ggsave(here('output', 'sp_range', '20220926_minmax-box_polychaetes.png'))

### Bivalves

# min and max for each species
## in points; one pair of points per species
ggplot(biv_rt) +
  geom_point(aes(x = reorder(index, min_lat), y = max_lat), 
             color = 'blue',
             alpha = .7) +
  geom_point(aes(x = reorder(index, min_lat), y = min_lat), 
             color = 'red',
             alpha = .4) +
  labs(title = 'Bivalve minimum and maximum occurrences', y = 'Latitude', x = 'Species') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_blank())

ggsave(here('output', 'sp_range', '20220926_minmax-dots_bivalves.png'))


# in boxplots
ggplot(biv_rt) + 
  geom_boxplot(aes(y = max_lat, x = 1), 
               fill = 'blue', 
               alpha = .5) +
  geom_boxplot(aes(y = min_lat, x = 2), 
               fill = 'red', 
               alpha = .5) +
  labs(title = 'Bivalve minimum and maximum occurrences', y = 'Latitude', x = '') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid = element_blank())

ggsave(here('output', 'sp_range', '20220926_minmax-box_bivalves.png'))
```




# Mapping

```{r prepare-to-map}
## Polychaetes
poly_clust2 <- poly_clust %>%
  group_by(cluster) %>%
  summarise(min_lat_clust = min(min), max_lat_clust = max(max))

poly2 <- poly %>%
  left_join(poly_rt) %>%
  filter(!(is.na(min_lat))) %>%
  mutate(in_cluster = ifelse((decimal_latitude >= 30 & 
                                decimal_latitude <= 32), 6, ifelse((decimal_latitude >= 32 & 
                                decimal_latitude <= 37), 5, ifelse((decimal_latitude >= 37 & 
                                decimal_latitude <= 49), 3, ifelse((decimal_latitude >= 49 & 
                                decimal_latitude <= 55), 9, ifelse((decimal_latitude >= 55 & 
                                decimal_latitude <= 60), 10, ifelse((decimal_latitude >= 60 & 
                                decimal_latitude <= 62), 11, NA_character_)))))))



## Bivalves

biv_clust2 <- biv_clust %>%
  group_by(cluster) %>%
  summarise(min_lat_clust = min(min), max_lat_clust = max(max))

biv2 <- biv %>%
  left_join(biv_rt) %>%
  filter(!(is.na(min_lat))) %>%
  mutate(in_cluster = ifelse((decimal_latitude >= 30 & 
                                decimal_latitude <= 32), 8, ifelse((decimal_latitude >= 32 & 
                                decimal_latitude <= 35), 7, ifelse((decimal_latitude >= 35 & 
                                decimal_latitude <= 48), 4, ifelse((decimal_latitude >= 48 & 
                                decimal_latitude <= 55), 5, ifelse((decimal_latitude >= 55 & 
                                decimal_latitude <= 60), 10, ifelse((decimal_latitude >= 60 & 
                                decimal_latitude <= 62), 11, NA_character_)))))))
```


## Visualize mCART results

### Polychaetes separately

Update with tree labeled and colored when internet

numbers on tree:
.9096 = 11
.7919 = 10
.6395 = 9
.1483 = 6
.7853 = 5
.2377 = 3
```{r}
plot(poly_tree)

```




```{r polychaete map}

ggplot() + geom_map(data = world, 
                    map = world, 
                    aes(x = long, y = lat, map_id = region), 
                    fill = "grey80") +
  geom_point(data = poly2, aes(x = decimal_longitude, y = decimal_latitude,
                               color = as.factor(in_cluster)),
             alpha = .5) +
  geom_point(data = poly_clust, aes(x = -137, y = min,  fill = as.factor(cluster)),
             shape = 23,
             size = 4) +
  scale_color_manual(values = poly_cluster_pal) +
  scale_fill_manual(values = poly_cluster_pal) +
    xlim(-142.5, -116.4) + 
  ylim(30, 62) +
  coord_fixed() +
  theme_bw() +
  labs(title = 'Polychaete clusters') +
  theme(axis.title = element_blank(),
        legend.position = 'none',
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

ggsave(here('output', 'maps', '20220927_cluster-map_polychaetes.png'),
       width = 6, height = 7)
```



## Bivalves


note: accidentally saved the same tree twice, rerun when I have internet.

```{r}
plot(biv_tree)
```



```{r}
## basic map
ggplot() + geom_map(data = world, 
                    map = world, 
                    aes(x = long, y = lat, map_id = region), 
                    fill = "grey80") +
  geom_point(data = biv2, aes(x = decimal_longitude, y = decimal_latitude,
                              color = as.factor(in_cluster)),
             alpha = .5) +
  geom_point(data = biv_clust, aes(x = -137, y = min,  fill = as.factor(cluster)),
             shape = 24,
             size = 4) +
  scale_color_manual(values = biv_cluster_pal) +
  scale_fill_manual(values = biv_cluster_pal) +
    xlim(-142.5, -116.4) + 
  ylim(30, 62) +
  coord_fixed() +
  labs(title = 'Bivalve Clusters') +
  theme_bw() +
  theme(axis.title = element_blank(),
        legend.position = 'none',
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

ggsave(here('output', 'maps', '20220927_cluster-map_bivalves.png'),
       width = 6)
```

## Together

```{r}

## basic map
ggplot() + geom_map(data = world, 
                    map = world, 
                    aes(x = long, y = lat, map_id = region), 
                    fill = "grey80") +
  geom_point(data = poly_clust, aes(x = -137, y = min,  
                                    fill = as.factor(cluster)),
             shape = 23,
             size = 4) +
  scale_fill_manual(values = poly_cluster_pal) +
  new_scale_fill()+
  geom_point(data = biv_clust, aes(x = -135, y = min,  
                                   fill = as.factor(cluster)),
             shape = 24,
             size = 4) +
  scale_fill_manual(values = biv_cluster_pal) +
    xlim(-140, -116.4) + 
  ylim(30, 62) +
  coord_fixed() +
  theme_bw() +
  theme(axis.title = element_blank(),
        legend.position = 'none',
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

ggsave(here('output', 'maps', '20220927_cluster-map_bins.png'),
       width = 6, height = 7)

```






```{r}
#this was the ggtree version

ggtree(coi_cons) %>%
  # move pelagic to the top
  flip(52, 32) %>% 
  flip(27, 31) %>%
  flip(29, 30) %<+%
  # add labels and coloring to ggtree
  coi_subset_plot +
  geom_tree(aes(color=as.factor(cluster))) +
  geom_text(aes(color = site, label= habitat_plot), 
            # slightly off the tips
            position = position_nudge(x = .005, y = 0),
            # left justify
            hjust=0, 
            size=3) +
  scale_colour_manual(values = site_pal) +
    # adding colors to branches AND tip labels means they get added to the color palette
  # need enough colors for all options
  #scale_color_manual(values = c('blue', 'black', 'grey40', 'green', 'orange',
  #                              'red', 'steelblue', 'skyblue', 'yellow', 'chartreuse',
  #                              'cyan', 'darkkhaki', 'darkorchid', 'aquamarine', 'deepskyblue',
  #                              'darkolivegreen3', 'deeppink1')) +
  # adjust xlim to fit text
  xlim(0, .5) +
  theme(legend.position = 'none')
```

